---
title: "Data Analysis & Visualization"
output: html_document
---


\
\

# Data Analysis

\


### 1. Load packages and cleaned dataset

* * *

```{r eval=FALSE}
library(tidyverse)
library(data.table)
library(ggplot2)
library(scales)
```

```{r include=FALSE}
library(tidyverse)
library(data.table)
library(ggplot2)
library(scales)
```


```{r eval=FALSE}
trips_cleaned <- read_csv("C:/Users/jains/Downloads/Cyclistic Project/Trip_Data_082021-072022/Cyclistic_Combined_Data.csv")
```

```{r include=FALSE}
trips_cleaned <- read_csv("C:/Users/jains/Downloads/Cyclistic Project/Trip_Data_082021-072022/Cyclistic_Combined_Data.csv")
```

\
\
\

### 2. Ride Length Based On various factors

* * *

**2.1 Average Trip Time Vs Membership Type**

We can check and compare the average trip time for both membership types.

```{r}
trips_cleaned %>% 
  group_by(member_casual) %>% 
  summarize(trip_time_by_membership = mean(trip_time))
```

This shows that casual riders tend to spend more time on rides in comparison to subscribers. Let's check some more factors to analyze the same.

\

**2.2 Average Trip Time Vs Day of Week**

```{r}
trips_cleaned %>% 
  group_by(day_of_week) %>% 
  summarize(trip_time_by_day = mean(trip_time))
```
We could see from the data that trip time is usually longer on weekends in comparison to other days. We will further classify this data on the basis of rider type. 
\

**2.3 Average Trip Time Vs Membership and Day of Week**

```{r}
trips_cleaned %>% 
  group_by(member_casual,day_of_week) %>% 
  summarize(trip_time_by_day = mean(trip_time))
```

We have already seen from section 2.1 and 2.2 that rides on weekends are usually longer and from the above data this is more clear that trip time is longer in weekends for both casual as well as members. But casual riders ride time is comparatively more than members.
\

**2.4 Count of Rides Vs Day of week**

```{r}
trips_cleaned %>% 
  group_by(day_of_week) %>% 
  tally()
```
The count of rides being more on weekends could be a factor for more trip time.
\

**2.5 Average Ride Length & Count of Rides Vs Membership & Day of Week**

```{r}
trips_cleaned %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(total_trips_by_day = n(), mean_rides = mean(trip_time)) %>% 
  arrange(member_casual, day_of_week)
```
As we can see that in the case of casual riders, the total trips are more on weekends hence the trip time is  more on the same days. But in the case of members that is not the case. The count of rides are almost equal in weekends and weekdays. Hence this could not give a clear picture about the trip time. Let's explore further.

\
\
\

## 3. Measuring summary statistics for ride length

* * *

Let's check for the mean, median, min. and max. value of trip time. 

```{r}

summary(trips_cleaned$trip_time)
```

This shows us that there is huge difference between mean trip time and max trip time. So, we can say that there are some of the rides which were greater than 1 day or 86400 secs.

```{r}
#Creating a dataframe for trips which are greater than 1 day

greater_than_1_day <- trips_cleaned %>% 
  filter(trip_time>=86400) %>% 
  mutate(cont_days=trip_time/86400) %>% 
  arrange(-cont_days)

#Creating a dataframe for trips which are less than 1 day

less_than_1_day <- trips_cleaned %>% 
  filter(trip_time<86400 & trip_time>300) %>% 
  mutate(trip_mins=trip_time/60) %>%
  group_by(member_casual,trip_mins) %>% 
  summarize(count=n()) %>% 
  arrange(-trip_mins) 

```

We will now visualize the data for further analysis.

\
\
\

# Data Visualization

\

## 1. Total no of rides by category type

* * *

We can calculate and visualize the total no of rides for each membership category

```{r}
#No. of trips made by rider type

pie <- trips_cleaned %>% 
  group_by(member_casual) %>% 
  summarise(count=n(),Percentage=n()/nrow(.)) %>% 
  ggplot(aes(x="", y=count, fill=member_casual))+
  geom_bar(width = 1, stat = "identity")+
  coord_polar("y", start=0) +
  guides(fill = guide_legend(title = "Membership")) +
  labs(title="% of Total Rides by Each Category")

#Create a blank theme

blank_theme <- theme_minimal()+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=14, face="bold")
  )

#Using pie chart to show the count 

pie + scale_fill_brewer("Blues") + blank_theme +
  theme(axis.text.x=element_blank())+
  geom_text(aes(y = count/2 + c(0, cumsum(count)[-length(count)]), 
                label = percent(1-Percentage)), size=5) 
```
  
So, the data we are using consists of 58% of the rides which are done by members and 42% of the rides which are done by casual riders.

\
\
\

## 2. No. of rides each month

* * *

We can check for the count of rides on a monthly basis for each of the membership category.

```{r}
#Total Rides Vs Month

trips_cleaned %>% 
  group_by(month_name,member_casual) %>% 
  tally() %>% 
  ggplot(aes(x=month_name, y=n, group=member_casual, color=member_casual)) + geom_line(size=1) +
  scale_x_discrete(limits = month.abb) +
  scale_y_continuous(expand = c(0, 0), limits=c(0,400000),labels = label_number(suffix = " K", scale = 1e-3)) +
  labs(x="Month",y="Count of Rides", title="Most Popular Time Of Year") +
  guides(color = guide_legend(title = "Membership"))

```
  
From the graph we could see that there is a sudden increase in the no. of rides between May to October for casual as well as member riders. This might be due to the weather and climate conditions. We could infer that customers use Cyclistic bikes more often in the summer season than in the winter season.

\
\
\

## 3. Most busy time of the day.

* * *

We will check for the most busy time of the day i.e. the time at which the count of rides is maximum. We will calculate this in Hrs.

```{r}
#Creating a separate dataframe for most busy time of Day

busy_time <- trips_cleaned %>% 
  mutate(Hour=as.character(format(started_at, "%H"))) %>% 
  group_by(Hour,member_casual) %>% 
  count(Hour)
  
 
busy_time %>% 
  ggplot(aes(x=Hour,y=n,group=member_casual, color=member_casual)) + 
  geom_line(size=1) + 
  labs(x="Hour of Day",y="Count of rides", title="Most Popular Time Of Day") +
  scale_y_continuous(expand = c(0, 0), limits=c(0,300000),labels = label_number(suffix = " K", scale = 1e-3)) +
  guides(color = guide_legend(title = "Membership"))

```
  
It can be observed that the time around 5 PM is the most busy time of the day. In the case of members the graph becomes steep at around 8 AM and 5 PM. By this we can assume that these people are using CYclistic bikes for going to work and coming back.  
In the case of casual riders the steep is observed only at 5PM, this could be due two both the reasons. Either they are using it for coming back from work or they might use it for leisure.
\
\
\

## 4. Total no. of Trips vs Day of week

* * *

Let's calculate the total no. of trips per day in a week. 

```{r}
#No of rides per day of a week

##Creating a dataframe for rides per day of week

rides_day_week <- trips_cleaned %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(total_trips_by_day = n(), mean_rides = mean(trip_time)) %>% 
  arrange(member_casual, day_of_week)

##Arranging the rides in order of days

rides_day_week$day_of_week <- factor(rides_day_week$day_of_week, levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))

##Plotting a line chart to visualize 

rides_day_week %>% 
  ggplot(aes(x = day_of_week, y = total_trips_by_day, group = member_casual, color=member_casual)) +
  geom_line(size=1) +
  scale_y_continuous(expand = c(0, 0), limits=c(0,500000),labels = label_number(suffix = " K", scale = 1e-3)) +
  labs(x="Day Of Week",y="Count of rides", title="Count Of Rides Vs Day Of Week") +
  guides(color = guide_legend(title = "Membership"))

```
  
From above, we can conclude that more no. of rides are being done by members in weekdays as compared to casual riders.  
Whereas more no. of rides are done by casual riders in weekends than members.

\
\
\

## 5. Percentage of Total Rides for each ride type

* * *

```{r}

trips_cleaned %>% 
  group_by(rideable_type, member_casual) %>% 
  summarize(count_ride_type = n(),Perc_of_total_rides=(n()/nrow(.))*100) %>% 
  ggplot(aes(x=member_casual, y=Perc_of_total_rides , fill=member_casual)) +
  geom_col(position="dodge", show.legend = FALSE) +
  facet_wrap(~ rideable_type) +
  scale_y_continuous(labels=function(x) paste0(x,"%")) +
  labs(x="Membership",y="% of Total Rides",title="% Of Total Rides For Each Ride Type & Membership ") 

```
  
This shows us that members usually prefer classic bikes whereas casual riders don't have a fixed preference.
 
\
\
\

## 6. Average ride length for each day of week

* * *

```{r}

trips_cleaned %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(total_trips_by_day = n(), average_ride_length = mean(trip_time)) %>% 
  arrange(member_casual, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = average_ride_length, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x="Day of Week",y="Average Trip Time", title="Average Ride Length Vs Day of Week") +
  guides(fill = guide_legend(title = "Membership"))

```

\
\
\

## 7. Relationship between count of rides with different parameters

* * *

We can check the relationship between percentage of total rides with respect to rider type, day of week and membership criteria. Let's check this.

```{r}

trips_cleaned %>% 
  group_by(rideable_type,member_casual,day_of_week) %>% 
  summarize(total_trips=n(),Percentage_total=(n()/nrow(.))*100) %>% 
  ggplot(aes(x=member_casual, y=Percentage_total, fill=member_casual)) +
  geom_col(position="dodge" , show.legend=FALSE) +
  facet_grid(rideable_type ~ day_of_week) +
  scale_y_continuous(labels=function(x) paste0(x,"%")) +
  theme(axis.text.x=element_text(angle=45)) +
  labs(x="Day of week / Membership",y="% of Total Rides / Ride Type") 

```

\
\
\

## 8. Top 30 stations for casual riders

* * *

We will now find out the top 30 stations for casual riders. Using this we can make an inference that which stations were used mainly as the starting point for casual riders.

```{r}
#We will create a separate dataframe for finding top 30 stations

##Filtering the mambership for casual riders and joining start and end stations

stations <- trips_cleaned %>% 
  filter(member_casual=='casual') %>% 
  select(start_station_name,end_station_name) %>% 
  unite(stations,c(start_station_name,end_station_name),sep="-") %>% 
  group_by(stations) %>% 
  summarize(Count=n())

##Ordering the no. of times a station is used in descending order

stations <- stations[order(-stations$Count),]

##Extracting top 30 stations

stations <- stations[1:30,]

##Again splitting joined station names to check for same start and end stations

stations[c('start_station_name', 'end_station_name')] <- str_split_fixed(stations$stations, '-', 2)

stations <- stations %>% 
  mutate(Same_station=ifelse(start_station_name==end_station_name,'Yes','No'))
```
\

We have now created a separate dataframe for all the stations with the no. of times they are being used. We found that some instances were present which had the same start and end station names. We will show them in our visualization.

\

```{r}

stations %>% 
  ggplot(aes(y=reorder(stations, +Count),x=Count, fill=Same_station)) + geom_col() +
  labs(x="Total Trips", y="Start Station - End Station", title="Top 30 Stations - Casual Riders") +
  guides(fill = guide_legend(title = "Same station"))

```
  
It can be observed that most of the rides done by casual riders have the same starting and ending station names. This could be the reason for longer trip time for casual riders as they tend to keep the bike with them for a longer time which results in increase of trip time.
\
\
\

## 9. Top 30 stations for member riders

* * *

After checking the top 30 stations for casual riders, it's time to check the same for member riders. Let's follow the same steps to check the same.

```{r}
#We will create a separate dataframe for finding top 30 stations

##Filtering the mambership for member riders and joining start and end stations

station_names_trips_members <- trips_cleaned %>% 
  filter(member_casual=='member') %>% 
  select(start_station_name,end_station_name) %>% 
  unite(stations,c(start_station_name,end_station_name),sep="-") %>% 
  group_by(stations) %>% 
  summarize(Count=n())

##Ordering the no. of times a station is used in descending order

station_names_trips_members <- station_names_trips_members[order(-station_names_trips_members$Count),]

##Extracting top 30 stations

station_names_trips_members <- station_names_trips_members[1:30,]

##Again splitting joined station names to check for same start and end stations

station_names_trips_members[c('start_station_name', 'end_station_name')] <- str_split_fixed(station_names_trips_members$stations, '-', 2)

station_names_trips_members <- station_names_trips_members %>% 
  mutate(Same_station=ifelse(start_station_name==end_station_name,'Yes','No'))

```

\

```{r}

station_names_trips_members %>% 
  ggplot(aes(y=reorder(stations, +Count),x=Count, fill=Same_station)) + geom_col() +
  labs(x="Total Trips", y="Start Station - End Station", title="Top 30 Stations - Members") +
  guides(fill = guide_legend(title = "Same station"))

```
  
In case of members the case is different as there are very few instances where starting and ending station names are the same. Hence, this can be concluded that members usually use Cyclistc bikes to commute to work.

\
\
\

## 10. Rides greater than 24 hours

* * *

Previously in section 3, we checked that there were some rides that were greater than 24 hours or 1 day. We will now visualize to check which rider type uses the Cyclistic bikes for more than 1 day.  
\
The dataframe created earlier for this was greater_than_1_day.

```{r}
#Scatterplot showing no. of rides greater than 1 day

greater_than_1_day %>% 
  ggplot(aes(x = 1:nrow(greater_than_1_day), y = cont_days)) +  
  geom_point(color="green") +
  facet_wrap(~ member_casual) + 
  labs(title="No. of rides greater than 1 Day", x="No. of rides", y="No. of Days")

```
  
From the above chart we can see that there are many instances in case of casual riders where the trip time is greater than 1 day with the highest trip being for almost 28-30 days while the maximum trip time in the case of members was around 1-2 days.

\
\
\

## 11. Rides less than 1 day

* * *

In the similar way, let's visualize the no. of rides less than 1 day. We will use trip time in mins to show this. The dataframe created earlier for this purpose was less_than_1_day.  
Please note that we are only considering the rides greater than 5 mins.

```{r}
#Scatterplot showing count of rides less than 1 day

less_than_1_day %>% 
  ggplot(aes(x = count, y = trip_mins)) +  
  geom_point(color="green") + 
  facet_wrap(~member_casual) +
  labs(title="No. of rides less than 1 Day", x="Count of rides", y="Trip Time(mins")

```

There were many rides for the members which were less than 10 mins. This might be because of the membership as they are using Cyclistic for small distances too. While in the case of casual riders small rides are less.
\
\
\